Parsing Postfix log files

### 2009-03-18 doc/logparser-architecture.tex
Removed some unused references.
Dealt with some XXX items, and added more.

\label{why separate rules, actions, and framework?}

@@

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua.  If an implementation is going
to use conditions so complex that they will require a Turing-equivalent
programming language to implement them, the design may need to be
re-thought, including the decision to use this architecture --- there may
be other architectures more suitable.

>>>>

@@

\label{rule conditions in architecture}

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua.  If an implementation is going
to use conditions so complex that they will require a programming language
capable of implementing a Turing machine, the design may need to be
re-thought, including the decision to use this architecture --- there may
be other architectures more suitable.

### 2009-03-18 doc/logparser-implementation.tex
Dealt with some XXX items, and added more.

    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
        lines the rule applies to.  This avoids needlessly trying rules
        that will not match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines that are not parsed by program specific
        rules.  This is a simple example of the rule conditions described
        in \sectionref{attaching conditions to rules}.

@@

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client: log the rejection
        with a mail if there is a queueid in the log line, or with the
        connection if not.  XXX VERY IMPORTANT ACTION\@; EXPAND AND
        REWRITE\@.

@@

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver.  This action is used for both \daemon{qmgr}
        and \daemon{cleanup} because it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup log lines} for details.
        XXX EXPLAIN THIS BETTER\@.

@@

XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@; IT
ALSO NEEDS TO BE PROOFREAD\@.

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes, they can take advantage of the parser's object oriented
implementation by sub-classing it and implementing their changes in the
derived class, allowing future upgrades of the parser with greatly reduced
chance of conflicts.\footnote{The real difference between the two
approaches is where the new code is placed.  The simpler option is to
change the parser code directly, but those changes will then have to be
made to subsequent versions of the parser, and as the scope of the changes
increases so does the chance of conflict or mistakes.  The second option
requires more effort initially but less effort in the long run: write a
subclass containing the new actions and change the program that invokes the
parser so that it uses the subclass rather than the parent class; the
changes required are minor and much less likely to lead to conflicts when
upgrading to future versions of the parser.  An alternative is to submit
new actions to the author \parsername{} for inclusion in future versions so
that the new actions do not need to be maintained separately, and other
users of the parser can benefit from the new functionality.} The action may
need to extend the list of valid combinations described in \sectionref{out
of order log lines} if the new action creates a different set of acceptable
programs, but this is unlikely to occur, as it would require parsing log
lines from Postfix components the parser currently ignores.\footnote{The
mail server used for development does not utilise either the \daemon{lmtp}
or \daemon{virtual} delivery agents, so this parser does not have rules to
handle log lines from those components.  Adding new rules to parse those
component's log lines is a simple process, though if their behaviour
differs significantly from the \daemon{smtp} or \daemon{local} delivery
agents new actions may be required.  The mail server in question is a
production mail server handling mail for a university department; the
benefit of using this server is that the log files used exhibit the
idiosyncrasies and peculiarities a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

XXX MENTION RULE CONDITIONS HERE\@; STRIP THE EXISTING STUFF FROM THE RULES
TABLE AND EXAMPLE RULE, THEN HAVE THEM REFER TO HERE\@.

@@

The theoretical aspects of this component are covered in detail in
\sectionref{rules in architecture}; this section is concerned with the
practical aspects of how rules are implemented and used in \parsername{}.
The structure of the rules table has already been documented in
\sectionref{rules table}; that abstract description will not be repeated
here, but should be fleshed out by the concrete example rule in
\sectionref{example rule in implementation}.  The process of creating new
rules from unparsed log lines is dealt with in \sectionref{creating new
rules in implementation}; the implementation of the utility supplied with
\parsername{} to create regexes from unparsed log lines is also described
in that section.  The final portion of this section provides some
suggestions for how to detect overlapping rules.

@@

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's regex will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.  Data captured by the regex (sender,
        recipient, etc.) will be automatically saved to the results and
        connections tables.

@@

@@

Every piece of software suffers from some limitations and there is almost
always room for improvement.  Below are the limitations and possible
improvements that have been identified in the parser.

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \acronym{DNSBL}, a new policy server, or new
        administrator-defined rejection messages requires new rules.

    \item The hostname used in the HELO command is not logged if the mail
        is accepted.  Rectifying this is relatively simple: create a
        restriction that is guaranteed to warn for every accepted mail, as
        follows:

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not be a queueid
        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
        evaluated for the first recipient, so the log line cannot be
        associated with the correct mail.  There is guaranteed to be a
        queueid when the DATA command has been reached, and thus the
        queueid will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}, and the log line can be
        associated with the correct mail.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

@@

    \item The parser does not separate mails where one or more mails are
        rejected and a subsequent mail is accepted; it will appear in the
        database as one mail with lots of rejections followed by acceptance
        (this has already been mentioned in \sectionref{connection reuse}).
        It does not appear to be possible to make this distinction given
        the data Postfix logs, though it might be possible to write a
        policy server to provide additional logging.

    \item The parser will not detect that it is parsing the same log file
        twice, resulting in the database containing duplicate entries.

    \item The parser does not distinguish between log files produced by
        different sources when parsing; all results will be saved to the
        same database.  This may be viewed as an advantage, as log files
        from different sources can be combined in the same database, or it
        may be viewed as a limitation as there is no facility to
        distinguish between log files from different sources in the same
        database.  If the results of parsing log files from different
        sources must remain separate, the parser can easily be instructed
        to store the results in a different database.

@@

    \item Improve the heuristics used in \sectionref{timeouts during data
        phase}, or develop another solution, to avoid incorrectly leaving a
        mail in the state tables.

>>>>

    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
        lines the rule applies to; see \sectionref{rule conditions in
        implementation} for full details of how this attribute is used.

@@

    \item [CLEANUP\_PROCESSING] \daemon{cleanup} processes every mail that
        passes through Postfix; details of what it does are available in
        \sectionref{Postfix Daemons}.  This action saves all data captured
        by the rule's regex if the log line has not come after a timeout
        (see \sectionref{discarding cleanup log lines}); it also creates
        the mail if necessary (see \sectionref{pickup logging after
        cleanup}).

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client.  This is the action
        most frequently specified by rules, because so many different
        restrictions are used to reject delivery attempts.  This action is
        quite simple: if the log line contains a queueid, save the data
        captured by the rule's regex to the mail identified by that
        queueid; otherwise save it to the connection identified by the
        \acronym{pid} in the log line.  No further processing of the log
        line is required.

@@

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver.  This action needs to deal with out of order
        log lines when mail is re-injected for forwarding; see
        \sectionref{Re-injected mails} for details.

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their new action.  The
action may need to extend the list of valid combinations described in
\sectionref{out of order log lines} if the new action creates a different
set of acceptable programs, but this would only be necessary if the new
action processes log lines from Postfix components \parsername{} currently
does not have rules for.

@@

@@

The theoretical aspects of this component are covered in detail in
\sectionref{rules in architecture}; this section is concerned with the
practical aspects of how rules are implemented and used in \parsername{}.
The structure of the rules table has already been documented in
\sectionref{rules table}; that abstract description will not be repeated
here, but should be fleshed out by the concrete example rule in
\sectionref{example rule in implementation}.  The process of creating new
rules from unparsed log lines is dealt with in \sectionref{creating new
rules in implementation}; the implementation of the utility supplied with
\parsername{} to create regexes from unparsed log lines is also described
in that section.  The rule conditions used in \parsername{} are explained
next, and the final portion of this section provides some
suggestions for how to detect overlapping rules.

@@

    \item [program and regex:] The program is used to restrict the log
        lines this rule will be used to recognise; see \sectionref{rule
        conditions in implementation} for details.  The regex does the
        actual recognition of the log lines, and data captured by the regex
        (sender, recipient, etc.) will be automatically saved to the
        results and connections tables.

@@

\subsection{Rule Conditions}

Rule conditions as part of the architecture have already been documented in
\sectionref{rule conditions in architecture}, and they can be very complex
and difficult to evaluate.  In contrast, \parsername{} uses very simple
rule conditions: each rule has a \texttt{program} attribute that specifies
the Postfix component whose log lines it recognises, and each log line
contains the name of the Postfix component that produced it; the framework
will only use rules where the two are equal when trying to recognise log
lines.  This avoids needlessly trying rules that will not match the log
line, or worse, might match unintentionally.  In addition the framework
supports generic rules, whose program attribute is ``*''; these will be
used if none of the program-specific rules recognise the log line.  If none
of the rules are successful the framework will warn the user.

\subsection{Regex Components}

XXX WRITE THIS BIT, AND REDUCE THE ENTRY IN THE RULES TABLE\@.

    The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of captured data; and makes each regex largely
        self-documenting.  For efficiency the keywords are expanded and
        every rule's regex is compiled before attempting to parse the log
        file; otherwise every regex would be recompiled each time it was
        used, resulting in a large, data dependent slowdown, as shown in
        \sectionref{Caching compiled regexes}.

@@

Every piece of software suffers from some limitations, and there is almost
always room for improvement.  Below are the limitations and possible
improvements that have been identified in \parsername{}.

@@

    \item Each new Postfix release requires writing new rules or modifying
        existing rules to cope with the new or changed log lines.
        Similarly using a new \acronym{DNSBL}, a new policy server, or new
        administrator-defined rejection messages requires new rules.

    \item The hostname used in the HELO command is not logged if the
        incoming delivery attempt is successful.  Configuring Postfix to do
        this is relatively simple: create a restriction that is guaranteed
        to warn for every accepted mail, as follows:

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not yet be a queueid
        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
        evaluated for the first recipient, so the log line cannot be
        associated with the correct mail.  There is guaranteed to be a
        queueid when the DATA command has been reached, and thus the
        queueid will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}, and the log line can be
        associated with the correct mail.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

@@

    \item \parsername{} does not create separate mails where one or more
        delivery attempts are rejected and subsequently a mail is accepted;
        it will appear in the database as one mail with lots of rejections
        followed by acceptance (this has already been mentioned in
        \sectionref{connection reuse}).  It does not appear to be possible
        to make this distinction given the data Postfix logs, though it
        might be possible to write a policy server to provide additional
        logging.

    \item \parsername{} will not detect that it is parsing the same log file
        twice, resulting in the database containing duplicate entries.

    \item \parsername{} does not distinguish between log files produced by
        different sources when parsing; all results will be saved to the
        same database.  This may be viewed as an advantage, because log
        files from different sources can be combined in the same database,
        or it may be viewed as a limitation because there is no facility to
        distinguish between log files from different sources in the same
        database.  If the results of parsing log files from different
        sources must remain separate, the parser can easily be instructed
        to use a different database.

@@

### 2009-03-18 doc/logparser-results.tex
Dealt with some XXX items, and added more.

\chapter{Results}

@@

XXX SHOULD THIS CHAPTER BE CALLED ``Results'' OR ``Evaluation''\@?

@@

XXX USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
ORDERINGS SECTION\@?

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
parse large log files.  The server that generated the log files used in
testing this parser accepts approximately 10,000 mails for 700 users a day;
\graphref{Mails received per day} shows that as expected there are far more
mails received on weekdays than at weekends.   Median log file size is
50MB, containing 285,000 log lines --- large scale mail servers would have
much larger log files.  Note that \graphref{Mails received per day} and
\tableref{Number of mails received per day: statistics} show the number of
mails received by \acronym{SMTP} only; in particular the mail loops
noticeable in other graphs do not contribute to these figures.

>>>>

\chapter{Results and Evaluation}

@@

@@

XXX USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
ORDERINGS SECTION\@?  MIX WITH RECOGNISING VS RECOGNISING + ACTION, AND
KEEP COUNTERS FOR EACH PROGRAM TOO --- MAYBE SOMETHING USEFUL WILL COME
FROM IT\@.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
parse large log files.  The server that generated the log files used in
testing this parser accepts approximately 10,000 mails for 700 users a day;
\graphref{Mails received per day} shows that as expected there are far more
mails received on weekdays than at weekends.  The mail server in question
is a production mail server handling mail for a university department; the
benefit of using this server is that the log files from it exhibit the
idiosyncrasies and peculiarities a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.  Median
log file size is 50MB, containing 285,000 log lines --- large scale mail
servers would have much larger log files.  Note that \graphref{Mails
received per day} and \tableref{Number of mails received per day:
statistics} show the number of mails received by \acronym{SMTP} only; in
particular the mail loops noticeable in other graphs do not contribute to
these figures.
